```{r}
install.packages("nlmeU")
install.packages("nlme")
install.packages("lattice")
install.packages("corrplot")
install.packages("plot.matrix")
install.packages("dplyr")
install.packages("lubridate")
```

```{r}
library(nlmeU)  # For dataset manipulation
library(nlme)   # For models implementation
library(lattice)
library(corrplot)
library(plot.matrix)
library(dplyr)
library(lubridate) # For date parsing
```

```{r}
# Load and inspect the data
data <- read.csv("daily_zones_data.csv")
date
```

```{r}
lm9.form <- formula(Price ~ -1 + Gas_Price + Month + Inflow + Outflow + Wind + Solar + Load + Decoupling + zone + zone:time_index)
lm9 <- lm(lm9.form, data = data)
summary(lm9)
```
```{r}
plot(fitted(lm9), residuals(lm9), main = "Residuals vs Fitted (lm9)", xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")
```
```{r}
library(ggplot2)  # For ggplot


# Set the size of the plotting area
par(pin = c(30, 10))

# Create plot_data dataframe
plot_data <- data.frame(
  fitted = data$Load,  # Use the fitted values from the gls model
  residuals = residuals(lm9),  # Use the residuals from the gls model
  c = data$Solar
)

# Plotting residuals vs fitted values
ggplot(plot_data, aes(x = fitted, y = residuals, color = c)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs Fitted",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
```


```{r}
plot(data$time_index , residuals(lm9), main = "Residuals vs Fitted (lm9)", xlab = "time_index", ylab = "Residuals")
abline(h = 0, col = "red")
```
```{r}
# Ensure that 'zone' is a factor
data$zone <- as.factor(data$zone)

# Plot: Residuals vs Time Index for all zones
plot(data$time_index, residuals(lm9), 
     main = "Residuals vs Time Index", 
     xlab = "Time Index", ylab = "Residuals",
     col = rgb(0, 0, 1, alpha = 0.6),  # Transparency with blue color
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

```


```{r}
# Ensure that 'zone' is a factor
data$zone <- as.factor(data$zone)

# Set up a 7x1 grid for 7 plots (7 rows, 1 column)
par(mfrow = c(1, 1), mar = c(3, 4, 2, 1))  # Reduce margins (bottom, left, top, right)

# Loop over the unique zones and create a plot for each zone
for (z in levels(data$zone)) {
  # Subset data for the current zone
  zone_data <- subset(data, zone == z)
  
  # Get the residuals for the current zone
  zone_residuals <- residuals(lm9)[data$zone == z]
  
  # Plot for each zone with transparency using rgb()
  plot(zone_data$time_index, zone_residuals, 
       main = paste("Residuals vs Time Index for Zone", z),
       xlab = "Time Index", ylab = "Residuals",
       col = rgb(0, 0, 1, alpha = 0.6),  # Apply transparency with rgb()
       pch = 1,  # Circle points
       cex = 1.2)  # Adjust size of points
  
  # Add horizontal line at y = 0
  abline(h = 0, col = "red")
}

# Reset the plotting area to default
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)  # Reset margins to default

```
```{r}
# Ensure that 'zone' is a factor
data$zone <- as.factor(data$zone)

# Plot: Residuals vs Time Index for all zones
plot(data$Wind, residuals(lm9), 
     main = "Residuals vs Time Index", 
     xlab = "gas", ylab = "Residuals",
     col = rgb(0, 0, 1, alpha = 0.6),  # Transparency with blue color
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")  
```
```{r}
# Ensure that 'zone' is a factor
data$zone <- as.factor(data$zone)

# Set up a 7x1 grid for 7 plots (7 rows, 1 column)
par(mfrow = c(1, 1), mar = c(3, 4, 2, 1))  # Reduce margins (bottom, left, top, right)

# Loop over the unique zones and create a plot for each zone
for (z in levels(data$zone)) {
  # Subset data for the current zone
  zone_data <- subset(data, zone == z)
  
  # Get the residuals for the current zone
  zone_residuals <- residuals(lm9)[data$zone == z]
  
  # Plot for each zone with transparency using rgb()
  plot(zone_data$Wind, zone_residuals, 
       main = paste("Residuals vs Time Index for Zone", z),
       xlab = "Time Index", ylab = "Residuals",
       col = rgb(0, 0, 1, alpha = 0.6),  # Apply transparency with rgb()
       pch = 1,  # Circle points
       cex = 1.2)  # Adjust size of points
  
  # Add horizontal line at y = 0
  abline(h = 0, col = "red")
}

# Reset the plotting area to default
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)  # Reset margins to default
```
```{r}
# Ensure that 'zone' is a factor
data$zone <- as.factor(data$zone)

# Plot: Residuals vs Time Index for all zones
plot(data$Load, residuals(lm9), 
     main = "Residuals vs load ", 
     xlab = "load", ylab = "Residuals",
     col = rgb(0, 0, 1, alpha = 0.6),  # Transparency with blue color
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")  
```

```{r}
# Ensure that 'zone' is a factor
data$zone <- as.factor(data$zone)

# Plot: Residuals vs Load Colored by Zone
plot(data$Load, residuals(lm9), 
     main = "Residuals vs Load (Colored by Zone)", 
     xlab = "Load", ylab = "Residuals",
     col = as.factor(data$zone),  # Color points by 'zone'
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate zones
legend("topright", legend = levels(data$zone), 
       col = 1:length(levels(data$zone)), pch = 16, title = "Zone")

```

```{r}
# Ensure that 'zone' is a factor
data$zone <- as.factor(data$zone)

# Plot: Residuals vs solar Colored by Zone
plot(data$Solar, residuals(lm9), 
     main = "Residuals vs solar (Colored by Zone)", 
     xlab = "", ylab = "Residuals",
     col = as.factor(data$zone),  # Color points by 'zone'
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate zones
legend("topright", legend = levels(data$zone), 
       col = 1:length(levels(data$zone)), pch = 16, title = "Zone")

```
```{r}
# Ensure that 'zone' is a factor
data$zone <- as.factor(data$zone)

# Plot: Residuals vs solar Colored by Zone
plot(data$Wind, residuals(lm9), 
     main = "Residuals vs wind (Colored by Zone)", 
     xlab = "", ylab = "Residuals",
     col = as.factor(data$zone),  # Color points by 'zone'
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate zones
legend("topright", legend = levels(data$zone), 
       col = 1:length(levels(data$zone)), pch = 16, title = "Zone")

```


#### create a distinction of north and south
```{r}
# Define the zones that represent "north"
north_zones <- c("CNOR", "NORD")

# Create the dummy variable for 'north'
data$north_dummy <- ifelse(data$zone %in% north_zones, 1, 0)

# View the updated dataset
head(data)

```


```{r}
# Ensure that 'zone' is a factor
data$north_dummy <- as.factor(data$north_dummy)

# Plot: Residuals vs Load Colored by Zone
plot(data$Solar, residuals(lm9), 
     main = "Residuals vs solar (Colored by Zone)", 
     xlab = "Load", ylab = "Residuals",
     col = as.factor(data$north_dummy),  # Color points by 'zone'
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate zones
legend("topright", legend = levels(data$north_dummy), 
       col = 1:length(levels(data$north_dummy)), pch = 16, title = "Zone")

```


```{r}
# Ensure that 'zone' is a factor
data$north_dummy <- as.factor(data$north_dummy)

# Plot: Residuals vs Load Colored by Zone
plot(data$Wind, residuals(lm9), 
     main = "Residuals vs solar (Colored by Zone)", 
     xlab = "Load", ylab = "Residuals",
     col = as.factor(data$north_dummy),  # Color points by 'zone'
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate zones
legend("topright", legend = levels(data$north_dummy), 
       col = 1:length(levels(data$north_dummy)), pch = 16, title = "Zone")

```

```{r}
# Ensure 'north_dummy' is a factor
data$north_dummy <- as.factor(data$north_dummy)

# Split the plotting area into two (1 row, 2 columns)
par(mfrow = c(1,1))

# Plot for north_dummy == 0
plot(data$Wind[data$north_dummy == "0"], 
     residuals(lm9)[data$north_dummy == "0"], 
     main = "Residuals vs Wind (north_dummy = 0)", 
     xlab = "Wind", ylab = "Residuals",
     col = "black", pch = 16, cex = 0.6)
abline(h = 0, col = "red")  # Add horizontal line at y = 0

# Plot for north_dummy == 1
plot(data$Wind[data$north_dummy == "1"], 
     residuals(lm9)[data$north_dummy == "1"], 
     main = "Residuals vs Wind (north_dummy = 1)", 
     xlab = "Wind", ylab = "Residuals",
     col = "pink", pch = 16, cex = 0.6)
abline(h = 0, col = "red")  # Add horizontal line at y = 0

# Reset the plotting area to default
par(mfrow = c(1, 1))

```
```{r}
# Ensure that 'zone' is a factor
data$north_dummy <- as.factor(data$north_dummy)

# Plot: Residuals vs Load Colored by Zone
plot(data$Inflow, residuals(lm9), 
     main = "Residuals vs inflow (Colored by Zone)", 
     xlab = "Load", ylab = "Residuals",
     col = as.factor(data$north_dummy),  # Color points by 'zone'
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate zones
legend("topright", legend = levels(data$north_dummy), 
       col = 1:length(levels(data$north_dummy)), pch = 16, title = "Zone")

```



```{r}
# Ensure that 'zone' is a factor
data$north_dummy <- as.factor(data$north_dummy)

# Plot: Residuals vs Load Colored by Zone
plot(data$Outflow, residuals(lm9), 
     main = "Residuals vs outflow (Colored by Zone)", 
     xlab = "Load", ylab = "Residuals",
     col = as.factor(data$north_dummy),  # Color points by 'zone'
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate zones
legend("topright", legend = levels(data$north_dummy), 
       col = 1:length(levels(data$north_dummy)), pch = 16, title = "Zone")

```


```{r}
# Ensure that 'zone' is a factor
data$north_dummy <- as.factor(data$north_dummy)

# Plot: Residuals vs Load Colored by Zone
plot(data$Load, residuals(lm9), 
     main = "Residuals vs Load (Colored by Zone)", 
     xlab = "Load", ylab = "Residuals",
     col = as.factor(data$north_dummy),  # Color points by 'zone'
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate zones
legend("topright", legend = levels(data$north_dummy), 
       col = 1:length(levels(data$north_dummy)), pch = 16, title = "Zone")

```

```{r}

# Ensure that 'zone' is a factor
data$north_dummy <- as.factor(data$north_dummy)

# Plot: Residuals vs Load Colored by Zone
plot(data$time_index, residuals(lm9), 
     main = "Residuals vs time (Colored by Zone)", 
     xlab = "time", ylab = "Residuals",
     col = as.factor(data$north_dummy),  # Color points by 'zone'
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate zones
legend("topright", legend = levels(data$north_dummy), 
       col = 1:length(levels(data$north_dummy)), pch = 16, title = "Zone")

```
```{r}
# Ensure 'time_index' is a POSIXct date-time object
data$datetime <- as.POSIXct(data$datetime)

# Extract the hour and create a new column 'hour'
data$hour <- as.numeric(format(data$datetime, "%H"))

# Ensure 'north_dummy' is a factor
data$north_dummy <- as.factor(data$north_dummy)

# Plot: Residuals vs Hour Colored by Zone
plot(data$hour, residuals(lm9), 
     main = "Residuals vs Hour (Colored by North Dummy)", 
     xlab = "Hour", ylab = "Residuals",
     col = as.factor(data$north_dummy),  # Color points by 'north_dummy'
     pch = 16,  # Circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate north_dummy
legend("topright", legend = levels(data$north_dummy), 
       col = 1:length(levels(data$north_dummy)), pch = 16, title = "North Dummy")


```
```{r}
# Ensure 'datetime' is a POSIXct date-time object
data$datetime <- as.POSIXct(data$date)

# Extract day of the week as numbers (0 = Sunday, 6 = Saturday) and create a new column 'day_of_week'
data$day_of_week <- as.numeric(format(data$datetime, "%w"))

# Ensure 'north_dummy' is a factor
data$north_dummy <- as.factor(data$north_dummy)

# Plot: Residuals vs Day of Week Colored by Zone
plot(data$day_of_week, residuals(lm9), 
     main = "Residuals vs Day of Week (Scatter Plot Colored by North Dummy)", 
     xlab = "Day of Week (0 = Sunday, 6 = Saturday)", 
     ylab = "Residuals",
     col = as.factor(data$north_dummy),  # Color points by 'north_dummy'
     pch = 16,  # Solid circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate north_dummy
legend("topright", legend = levels(data$north_dummy), 
       col = 1:length(levels(data$north_dummy)), pch = 16, title = "North Dummy")

```

```{r}
# Ensure 'datetime' is a POSIXct date-time object
data$datetime <- as.POSIXct(data$date)

# Extract day of the month (1–31) and create a new column 'day_of_month'
data$day_of_month <- as.numeric(format(data$datetime, "%d"))

# Ensure 'north_dummy' is a factor
data$north_dummy <- as.factor(data$north_dummy)

# Plot: Residuals vs Day of Month Colored by Zone
plot(data$day_of_month, residuals(lm9), 
     main = "Residuals vs Day of Month (Scatter Plot Colored by North Dummy)", 
     xlab = "Day of Month (1–31)", 
     ylab = "Residuals",
     col = as.factor(data$north_dummy),  # Color points by 'north_dummy'
     pch = 16,  # Solid circle markers
     cex = 0.6)  # Adjust size of points

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate north_dummy
legend("topright", legend = levels(data$north_dummy), 
       col = 1:length(levels(data$north_dummy)), pch = 16, title = "North Dummy")

```

```{r}
# Ensure 'datetime' is a POSIXct date-time object
data$datetime <- as.POSIXct(data$date)

# Extract month and create a 'season' column
data$season <- cut(as.numeric(format(data$datetime, "%m")),
                   breaks = c(0, 2, 5, 8, 11, 12),  # Define season boundaries
                   labels = c("Winter", "Spring", "Summer", "Fall", "Winter"),
                   include.lowest = TRUE)

# Assign a numeric value to seasons for plotting
data$season_num <- as.numeric(data$season)

# Ensure 'north_dummy' is a factor
data$north_dummy <- as.factor(data$north_dummy)

# Scatter Plot: Residuals vs Season (Colored by Zone)
plot(data$season_num, residuals(lm9), 
     main = "Residuals vs Season (Scatter Plot Colored by North Dummy)", 
     xlab = "Season (1 = Winter, 2 = Spring, 3 = Summer, 4 = Fall)", 
     ylab = "Residuals",
     col = as.numeric(data$north_dummy),  # Color points by 'north_dummy'
     pch = 16,  # Solid circle markers
     cex = 0.6,  # Adjust size of points
     xaxt = "n")  # Remove default x-axis ticks

# Add custom x-axis labels
axis(1, at = 1:4, labels = c("Winter", "Spring", "Summer", "Fall"))

# Add horizontal line at y = 0
abline(h = 0, col = "red")

# Add a legend to indicate north_dummy
legend("topright", legend = levels(data$north_dummy), 
       col = 1:length(levels(data$north_dummy)), pch = 16, title = "North Dummy")

```

```{r}
library(nlmeU)  # --> for the dataset
library(nlme)   # --> for models implementation
library(lattice)
library(corrplot)
library(plot.matrix)
```

#### Checking structures 
Variance structure for the reversed funnel shape in residuals vs solar:
The reversed funnel shape in the residuals vs solar plot indicates heteroscedasticity, where the variance of residuals changes with the level of solar output. To address this, you can use a variance structure that allows for different variances at different levels of solar output. A power variance function could be appropriate here.

Suggestion: Use the nlme package to fit a model with a power variance function for solar.

```{r}

# Assuming your original model is called lm9
gls_model <- gls(lm9.form, 
                 data = data, 
                 weights = varPower(form = ~Load), 
                 method = "REML")


summary(gls_model)
```

```{r}
library(ggplot2)  # For ggplot


# Set the size of the plotting area
par(pin = c(30, 10))

# Create plot_data dataframe
plot_data <- data.frame(
  fitted = fitted(gls_model),  # Use the fitted values from the gls model
  residuals = residuals(gls_model),  # Use the residuals from the gls model
  zone = data$zone  
)

# Plotting residuals vs fitted values
ggplot(plot_data, aes(x = fitted, y = residuals, color = zone)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs Fitted (GLS Model)",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
```

```{r}

# Create the plot
plot(data$time_index, residuals(gls_model), 
     main = "Residuals vs Fitted (gls)", 
     xlab = "time_index", 
     ylab = "Residuals")
abline(h = 0, col = "red")
```


```{r}
# varPower with different delta parameters per zone
fm6 <- gls(lm9.form, data = data, method = "REML", weights = varPower(form = ~ Gas_Price|zone ))
summary(fm6)
```
```{r}
# Extract residuals and fitted values
residuals <- residuals(fm6)  # Pearson residuals for better interpretation
fitted_values <- fitted(fm6)

# Create a residuals vs. fitted values plot
ggplot(data = data.frame(Fitted = fitted_values, Residuals = residuals), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "blue", alpha = 0.6) +  # Scatter points
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Horizontal line at zero
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

```









```{r}
# Define the autoregressive model for fm6 with AR(1) structure
fm6_AR1 <- gls(
  lm9.form,                                       # Model formula
  data = data,                                    # Dataset
  method = "REML",                                # Restricted maximum likelihood
  weights = varPower(form = ~ time_index | zone),       # Variance structure with varPower
  correlation = corAR1(form = ~time_index | zone)  # Autoregressive correlation structure within each zone
)

# Print the summary of the model
summary(fm6_AR1)
```

